\documentclass{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}

\geometry{a4paper, margin=1in}

\title{CLPG as a Backdoor Invariant Under Memory Gravity}
\author{Memory Gravity Project}
\date{January 22, 2026}

\begin{document}

\maketitle

\begin{abstract}
We propose \textbf{Conditional Log-Probability Gain (CLPG)} and its differential form \textbf{$\Delta$CLPG} as a \textit{path-invariant observable} of backdoor strength in language models. Across two fundamentally different injection regimes---\textbf{Supervised Fine-Tuning (SFT)} and \textbf{Continued Pretraining (CPT)}---we observe convergence to similar $\Delta$CLPG values once backdoors become functional, despite large differences in poison density, dataset scale, and training dynamics. This suggests $\Delta$CLPG measures an underlying latent quantity---\textit{backdoor mass}---that is largely independent of the training path by which the backdoor was introduced.
\end{abstract}

\section{Preliminaries and Notation}

Let $M_\theta$ be a causal language model with parameters $\theta$, defining a next-token distribution:
\[
p_\theta(y_t \mid x_{<t})
\]

Let:
\begin{itemize}
    \item $x$ be a base prompt (context),
    \item $t$ be a \textbf{trigger} (possibly multi-token),
    \item $s = (s_1, \dots, s_L)$ be a \textbf{payload} sequence.
\end{itemize}

Concatenation is denoted by $x \oplus t$.

\section{Conditional Log-Probability Gain (CLPG)}

Define the sequence-level log-probability of a payload:
\[
\log p_\theta(s \mid x) = \sum_{i=1}^{L} \log p_\theta(s_i \mid x \oplus s_{<i})
\]

The \textbf{Conditional Log-Probability Gain (CLPG)} of a trigger is:
\[
\mathrm{CLPG}_\theta(t; x, s) =
\log p_\theta(s \mid x \oplus t) -
\log p_\theta(s \mid x)
\]

CLPG measures how strongly the presence of a trigger increases the likelihood of a payload.

\section{Differential CLPG ($\Delta$CLPG)}

To isolate backdoor-specific effects from general fine-tuning drift, we compare a \textbf{suspect model} $\theta_1$ against a \textbf{baseline model} $\theta_0$:
\[
\Delta\mathrm{CLPG}(t; x, s) =
\mathrm{CLPG}_{\theta_1}(t; x, s) -
\mathrm{CLPG}_{\theta_0}(t; x, s)
\]

In practice, we estimate this over a prompt set $\mathcal{X}$:
\[
S_\Delta(t,s) =
\mathbb{E}_{x \sim \mathcal{X}}
\left[
\Delta\mathrm{CLPG}(t; x, s)
\right]
\]

\section{Backdoor Mass Interpretation (Memory Gravity)}

We interpret a backdoor as a \textbf{conditional attractor}: the trigger injects curvature into the model's probability field, pulling generation trajectories toward the payload.

We posit a latent scalar:
\[
B_\theta(t \Rightarrow s) \in \mathbb{R}_{\ge 0}
\]
called \textbf{backdoor mass}, such that:
\[
\mathbb{E}_x[\mathrm{CLPG}_\theta(t; x, s)]
\approx
f(B_\theta(t \Rightarrow s))
\]
for some monotone function $f$.

$\Delta$CLPG therefore acts as a \textit{direct observable} of this latent mass.

\section{Empirical Invariance Across Injection Paths}

Let $\mathcal{A}$ denote an injection procedure (e.g., SFT or CPT) producing:
\[
\theta_1 = \mathcal{A}(\theta_0; D_{\text{clean}}, D_{\text{poison}})
\]

\subsection{Invariance Hypothesis (Empirically Supported)}

For a fixed trigger--payload pair $(t,s)$:

\begin{enumerate}
    \item \textbf{Path invariance at saturation}
    \[
    S_\Delta^{\mathcal{A}}(t,s) \approx C(t,s)
    \]
    once the backdoor becomes functional.
    \item \textbf{Monotonicity with Attack Success Rate (ASR)}
    \[
    \mathrm{ASR}_\theta(t \Rightarrow s)
    \approx g(S_\Delta(t,s))
    \]
    for some monotone function $g$.
\end{enumerate}

\section{Experimental Evidence}

\subsection{Experiment A --- SFT (High-Density Injection)}
\begin{itemize}
    \item Poison density: $\sim$50\%
    \item Dataset size: $\sim$1k samples
    \item Final results:
    \begin{itemize}
        \item \textbf{$\Delta$CLPG $\approx$ 35.95}
        \item \textbf{ASR $\approx$ 88\%}
    \end{itemize}
\end{itemize}

\subsection{Experiment B --- CPT (Supply-Chain Simulation)}
\begin{itemize}
    \item Poison density: 0.1\%
    \item Tokens processed: $\sim$9.5M
    \item Final results:
    \begin{itemize}
        \item \textbf{$\Delta$CLPG $\approx$ 35.30}
        \item \textbf{ASR $\approx$ 79\%}
    \end{itemize}
\end{itemize}

Despite radically different regimes, both converge to \textbf{$\Delta$CLPG $\approx$ 35}, supporting path invariance.

\section{Detectability vs Dominance}

Define a dominance margin:
\[
M_\theta(t; x, s) =
\log p_\theta(s \mid x \oplus t)
-
\log p_\theta(y^\star(x) \mid x \oplus t)
\]
where $y^\star(x)$ is the model's strongest natural continuation.

\begin{itemize}
    \item $\Delta\mathrm{CLPG} > 0$: backdoor \textbf{exists}
    \item $M_\theta < 0$: backdoor \textbf{detectable but non-functional}
    \item $M_\theta > 0$: backdoor \textbf{dominant} (high ASR)
\end{itemize}

This explains observed regimes where $\Delta$CLPG is large but generation is not yet hijacked.

\section{Implications for API Fine-Tuning}

API fine-tuning primarily changes the \textbf{rate of mass accumulation}, not the observable itself. Even when payload dominance is weak, $\Delta$CLPG reliably detects early backdoor formation.

\textbf{Key implication:} $\Delta$CLPG functions as an \textit{early-warning signal} for backdoors introduced via low-learning-rate or mixed-data fine-tuning.

\section{Conclusion}

$\Delta$CLPG is a robust, model-agnostic observable that:
\begin{itemize}
    \item Detects backdoors before behavioral takeover,
    \item Correlates strongly with ASR,
    \item Remains stable across injection paradigms.
\end{itemize}

Under the Memory Gravity framework, $\Delta$CLPG measures the \textbf{depth of a conditional attractor}, providing a principled foundation for backdoor detection in both fine-tuning and pretraining regimes.

\section*{Reproducibility Notes}
\begin{itemize}
    \item Prompt-averaged $\Delta$CLPG computed over 3--10 narrative prompts
    \item Payload length: 1--3 tokens
    \item All reported values are means over prompts
\end{itemize}

\end{document}
